---
title: "EDAV Fall 2019 - Racial Segregation in NYC"
author: "Dhananjay Deshpande, Meenakshi Zutshi"
output: html_document
code_folding: hide
---

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
```

## Introduction

New York City is amongst the most populated cities in the world with about 8.6 million people (2017). It is also the melting pot of America with representation from all races.

We came across many articles speaking about racial segregation in NYC. https://www.nytimes.com/2019/07/16/nyregion/segregation-nyc-affordable-housing.html refers to how the city’s policy of giving preference to local residents for new affordable housing helps perpetuate racial segregation. As per https://www.nytimes.com/2019/03/26/nyregion/school-segregation-new-york.html, NYC public schools are still struggling with racial segregation.  

We were curious to look at demographic data to find out the racial distribution of population in New York City to help answer the following questions:

1. Are races segregated across NYC?
2. How are races segregated in New York City by boroughs and zip codes?
3. Is segregation of races consistent with racial segregation in schools?

## Datasets

We found multiple datasets are available for demographic and race data. We chose to focus on the following datasets :

1. US Government Census data includes data uptil 2018 regarding race and ethnicity by county and zipcode

- American Community Survey

https://data.census.gov/cedsci/profile?g=0400000US36&q=New%20York

Census data is collected every 10 years (decennial survey). Last survey was conducted in 2010.

The other alternative survey peformed by sampling the population is the American Community Survey (ACS), performed yearly. The datasets available from ACS are either yearly or average of past 5 years. The 5 year average is considered the most reliable. We choose to use the 2013-2017 5 year average data for our study.

The ACS data contains population numbers for the following races 

- American Indian
- Asian
- Black
- White
- Other Race (Used by Hispanics)
- Pacific Islander
- Two or more races

2. DOE New York City schools demographic data between 2013-2018 

- https://data.cityofnewyork.us/Education/2013-2018-Demographic-Snapshot-School/s52a-8aq6

## Analysis

```{r}
#install.packages("devtools")
#install_github('arilamstein/choroplethrZip@v1.5.0')
#install.packages("choroplethrMaps")
#install.packages("acs")
#install.packages("xlsx")
#install.packages("tidycensus")
#install.packages("janitor")
```


### Distribution of Races by Borough

We found choroplethr packages to map population by demographics to boroughs and zipcodes on a choropleth map of NYC. 

Choropleth maps allow us to make comparisons before data corresponding to geographical areas spatially. The borough maps were useful to get a sense of where each race was concentrated at the Borough level.

```{r}
# Choropleth maps for county and zip by race 

library(choroplethr)
library(acs)
library(ggplot2)
library(RColorBrewer)
library(dplyr)
library(choroplethrMaps)
library(choroplethrZip)

# Need to go to http://api.census.gov/data/key_signup.html to set API key
api.key.install("5090bf8957911bad76f58e6637e951fe219d5e45")

# Race by borough
nyc_fips = c(36005, 36047, 36061, 36081, 36085)
county_choropleth_acs(tableId="B01003", endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="Total NYC Population by County (2017)")
county_choropleth_acs(tableId="B02008",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by County (2017) - White")
county_choropleth_acs(tableId="B02009",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by County (2017) - Black ")
county_choropleth_acs(tableId="B02010",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by County (2017) - American Indian ")
county_choropleth_acs(tableId="B02011",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by County (2017) - Asian")
county_choropleth_acs(tableId="B02012",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by County (2017) - Pacific Islander")
county_choropleth_acs(tableId="B02013",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by County (2017) - Other Race")
```

**Observations from Borough Choropleth maps:**

 - population of White people is highest in Brooklyn than other boroughs
 - population of Black people is highest in Brooklyn than other boroughs
 - population of Asians people is highest in Queens than other boroughs
 - population of Hispanic people is highest in Bronx than other boroughs
 
### Distribution of Races by Zip Code

Choroplethr maps by zip code allowed us to identify clusters of zip codes where each race was concentrated.
 
```{r}
# Race by zipcode
zip_choropleth_acs(tableId="B01003", endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="Total NYC Population by Zipcode (2017)")
zip_choropleth_acs(tableId="B02008",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by Zipcode (2017) - White")
zip_choropleth_acs(tableId="B02009",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by Zipcode (2017) - Black ")
zip_choropleth_acs(tableId="B02010",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by Zipcode (2017) - American Indian")
zip_choropleth_acs(tableId="B02011",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by Zipcode (2017) - Asian")
zip_choropleth_acs(tableId="B02012",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by Zipcode (2017) - Pacific Islander")
zip_choropleth_acs(tableId="B02013",endyear = 2017, span=5, county_zoom=nyc_fips)+labs(title="NYC Population by Zipcode (2017) - Other Race")

```

**Observations from Choropleth maps By Zipcodes:**

 - High concentration of white population can be noticed in neighborhoods in lower Brooklyn, Central Staten Island and Upper West Side
 - High concentration of black population can be noticed in neighborhoods in southern Brookyln, Southern Queens
 - High concentration of Hispanic of populated can be noticed in neighborhoods in Bronx
 
### Studying Racial concentration by Zipcodes using Bar Charts
 
We noticed that Zipcodes in NYC have unique patterns per Borough:

- 100** for Manhattan (with the exception of 10280)
- 103** for Staten Island
- 104** for Bronx
- 112* for Brooklyn
- 11*** for Queens (except 112*)

We can get a sense of racial concentration by borough by plotting the populations per race against the sorted list of Zip Codes.

```{r}
library(tidycensus)
census <- load_variables(2017,"acs5",cache=TRUE)
```
```{r}
ny <- get_acs(geography = "zcta", 
              variables = c(total = "B02001_001", white = "B02001_002", black = "B02001_003", americanindian = "B02001_004",asian =  "B02001_005", pacificislander = "B02001_006", hispanic = "B03002_001", otherrace = "B02001_007", twoormoreraces = "B02001_008"), output="wide")
```


```{r}
library(ggplot2)
library(tidyverse)

mapping <- read.csv("New_York_State_ZIP_Codes-County_FIPS_Cross-Reference.csv")
nyc_mapping <- mapping[mapping$County.FIPS %in% nyc_fips,]

nyc_acs <-ny[ny$GEOID %in% nyc_mapping$ZIP.Code,]

nyc_acs <- nyc_acs[order(nyc_acs$GEOID),] 

```

```{r}
# Bar charts per race by zip code

library(ggplot2)
library(tidyverse)

total<-ggplot(nyc_acs, aes(x=nyc_acs$GEOID, y=nyc_acs$totalE)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 3)) +
  xlab("Zip codes") +
  ylab("Population")
    
total

whites<-ggplot(nyc_acs, aes(x=nyc_acs$GEOID, y=nyc_acs$whiteE)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=3)) +
  xlab("Zip codes") +
  ylab("Population")

whites

blacks<-ggplot(nyc_acs, aes(x=nyc_acs$GEOID, y=nyc_acs$blackE)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=3)) +
  xlab("Zip codes") +
  ylab("Population")

blacks

americanindians<-ggplot(nyc_acs, aes(x=nyc_acs$GEOID, y=nyc_acs$americanindianE)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=3)) +
  xlab("Zip codes") +
  ylab("Population")

americanindians

asians<-ggplot(nyc_acs, aes(x=nyc_acs$GEOID, y=nyc_acs$asianE)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=3)) +
  xlab("Zip codes") +
  ylab("Population")

asians

pacificislanders<-ggplot(nyc_acs, aes(x=nyc_acs$GEOID, y=nyc_acs$pacificislanderE)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=3)) +
  xlab("Zip codes") +
  ylab("Population")

pacificislanders

otherraces<-ggplot(nyc_acs, aes(x=nyc_acs$GEOID, y=nyc_acs$otherraceE)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=3)) +
  xlab("Zip codes") +
  ylab("Population")

otherraces

twoormoreraces<-ggplot(nyc_acs, aes(x=nyc_acs$GEOID, y=nyc_acs$twoormoreracesE)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=3)) +
  xlab("Zip codes") +
  ylab("Population")

twoormoreraces
```

**Observations from Bar Plots Per Zipcode for Individual Races:**

 - High concentration of whites in Brooklyn
 - High concentration of blacks in Brooklyn yet sparser than whites

### Comparing Races

Next we compared the populations between races across all the zip codes. This was to identify if there was a larger concentration of any one race wrt others across the zipcodes in NYC.

```{r}

# Plot representation by zipcodes

library(ggplot2)
library(tidyverse)

tidy_nyc_acs <- nyc_acs %>% 
  gather(Race, value, c(whiteE,blackE,americanindianE, asianE, pacificislanderE, otherraceE,twoormoreracesE))

ggplot(tidy_nyc_acs, aes(x = GEOID, y = value, fill = Race)) + 
  geom_col(position = position_stack()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=3)) +
  xlab("Zip Codes") + 
  ylab("Population")

```

**Observations from Stacked Bar Plots Per Zipcode:**

- There are zipcodes with predominantly white population (10128, 11211, 11385)
- There are zipcodes with predominantly black population (10466, 11236)
- There are zipcodes with predominantly hispanic population (10468)

```{r}
# Plot representation by neighborhoods
library(janitor)
library(tidyverse)

chartRegion <- function(zipCodeList, neighborhood, county)
{

  region <- ny[ny$GEOID %in% zipCodeList, ]



  region<- region[c("GEOID","whiteE","blackE","americanindianE", "asianE", "pacificislanderE", "otherraceE","twoormoreracesE")]

  region <- region %>%
  adorn_totals("row")

  n <- nrow(region)
  region <- region[(n):n, ]

  tidy_region <- region %>% 
  gather(Race, Population, -GEOID)

  ggplot(tidy_region, aes(x = GEOID, y = Population, fill = Race)) + 
  geom_col(position = position_stack()) +
  ggtitle(paste("Population in ", neighborhood, " - ", county)) +
  xlab(neighborhood)

}
```

### Comparing Racial Population by Neighborhood

Next we studied each neighborhood in NYC as per neighborhoods listed here - https://www.health.ny.gov/statistics/cancer/registry/appendix/neighborhoods.htm

The goal was to find neighborhoods where populations from one race dominated. 

#### Bronx
```{r}
# Bronx

chartRegion(c(10453, 10457, 10460), "Central Bronx", "Bronx")
chartRegion(c(10458, 10467, 10468), "Bronx Park and Fordham", "Bronx")
chartRegion(c(10451, 10452, 10456), "High Bridge and Morrisania", "Bronx")
chartRegion(c(10454, 10455, 10459, 10474), "Hunts Point and Mott Haven", "Bronx")
chartRegion(c(10463, 10471), "Kingsbridge and Riverdale", "Bronx")
chartRegion(c(10466, 10469, 10470, 10475), "Northeast Bronx", "Bronx")
chartRegion(c(10461, 10462,10464, 10465, 10472, 10473), "Southeast Bronx", "Bronx")
```

**Observations about Neighborhoods in Bronx:**

- Neighborhoods in Brnx wre found to have high concentration of Hispanic population compared to other races. Central Bronx, Bronx Park and Fordham, Hunts Point and Mott Haven had the highest concentration of Hispanic population.

#### Brooklyn

```{r}
# Brooklyn
par(mfrow = c(3, 4))
chartRegion(c(11212, 11213, 11216, 11233, 11238), "Central Brooklyn", "Brooklyn")
chartRegion(c(11209, 11214, 11228), "Southwest Brooklyn", "Brooklyn")
chartRegion(c(11204, 11218, 11219, 11230), "Borough Park", "Brooklyn")
chartRegion(c(11234, 11236, 11239), "Canarsie and Flatlands", "Brooklyn")
chartRegion(c(11223, 11224, 11229, 11235), "Southern Brooklyn", "Brooklyn")
chartRegion(c(11201, 11205, 11215, 11217, 11231), "Northwest Brooklyn", "Brooklyn")
chartRegion(c(11203, 11210, 11225, 11226), "Flatbush", "Brooklyn")
chartRegion(c(11207, 11208), "East New York and New Lots", "Brooklyn")
chartRegion(c(11211, 11222), "Greenpoint", "Brooklyn")
chartRegion(c(11220, 11232), "Sunset Park", "Brooklyn")
chartRegion(c(11206, 11221, 11237), "Bushwick and Williamsburg", "Brooklyn")
```

**Observations about Neighborhoods in Brooklyn:**

 - Central Brooklyn, Canarsie and Flatlands, Flatbush, East New York and New Lots are predominantly Black
 - Southwest Brooklyn, Borough Park, Southern Brooklyn, Northwest Brooklyn are predominantly White
 
#### Manhattan 
```{r}
# Manhattan

chartRegion(c(10026, 10027, 10030, 10037, 10039), "Central Harlem", "Manhattan")
chartRegion(c(10001, 10011, 10018, 10019, 10020, 10036), "Chelsea and Clinton", "Manhattan")
chartRegion(c(10029, 10035), "East Harlem", "Manhattan")
chartRegion(c(10010, 10016, 10017, 10022), "Gramercy Park and Murray Hill", "Manhattan")
chartRegion(c(10012, 10013, 10014), "Greenwich Village and Soho", "Manhattan")
chartRegion(c(10004, 10005, 10006, 10007, 10038, 10280), "Lower Manhattan", "Manhattan")
chartRegion(c(10002, 10003, 10009), "Lower East Side", "Manhattan")
chartRegion(c(10002, 10003, 10009), "Upper East Side", "Manhattan")
chartRegion(c(10023, 10024, 10025), "Upper West Side", "Manhattan")
chartRegion(c(10031, 10032, 10033, 10034, 10040), "Inwood and Washington Heights", "Manhattan")
```
**Observations about Neighborhoods in Manhattan:**

 - Central Harlem is predominantly Black
 - Chelsea and Clinton, Gramercy Park abd Murray Hill, Greenwich Village and Soho, Lower Manhattan, Upper West Side is predominantly White

```{r}
# Queens

chartRegion(c(11361, 11362, 11363, 11364), "Northeast Queens", "Queens")
chartRegion(c(11354, 11355, 11356, 11357, 11358, 11359, 11360), "North Queens", "Queens")
chartRegion(c(11365, 11366, 11367), "Central Queens", "Queens")
chartRegion(c(11412, 11423, 11432, 11433, 11434, 11435, 11436), "Jamaica", "Queens")
chartRegion(c(11101, 11102, 11103, 11104, 11105, 11106), "Northwest Queens", "Queens")
chartRegion(c(11374, 11375, 11379, 11385), "West Central Queens", "Queens")
chartRegion(c(11691, 11692, 11693, 11694, 11695, 11697), "Rockaways", "Queens")
chartRegion(c(11004, 11005, 11411, 11413, 11422, 11426, 11427, 11428, 11429), "Southeast Queens", "Queens")
chartRegion(c(11414, 11415, 11416, 11417, 11418, 11419, 11420, 11421), "Southwest Queens", "Queens")
chartRegion(c(11368, 11369, 11370, 11372, 11373, 11377, 11378), "West Queens", "Queens")
```

**Observations about Neighborhoods in Queens:**

 - West Central, Northwest Queens has much higher number of whites than other races
 - Jamaica and South East Queens has a much higher number of blacks than other races

#### Staten Island

```{r}
# Staten Island

chartRegion(c(10302, 10303, 10310), "Port Richmond", "Staten Island")
chartRegion(c(10306, 10307, 10308, 10309, 10312), "South Shore", "Staten Island")
chartRegion(c(10301, 10304, 10305), "Stapleton and St. George", "Staten Island")
chartRegion(c(10314), "Mid-Island", "Staten Island")

```
**Observations about Neighborhoods in Staten Island:**

- Staten Island in general has a higher number of whites than other races
- South Shore of Staten Island has very high concentration of whites compared to other races

## Conclusion:

- We were able to identify regions in NYC where there were very high concentrations of particular races in comparison to others
- Overall all areas have some representation from atleast 2 races as we did not find a single zip code with representation from only one race

## github link:

https://github.com/dcd2139/edav_final_project

## References:
 
https://dlab.berkeley.edu/sites/default/files/training_materials/Census_Lecture_030915.pdf
https://rdrr.io/cran/choroplethr/man/county_choropleth_acs.html
https://arilamstein.com/creating-zip-code-choropleths-choroplethrzip/
https://www.trulia.com/blog/tech/the-choroplethr-package-for-r/